Technical Report

### 2017-12-20 fully-abstract-comp/technical-report.tex
more small changes to the section about incstruction interpretation.

  \caption{Permission hierarchy}
  \label{fig:perm-hier}
\end{figure}
\lau{13-09-2017: Note that most of the discussion belove is now captured in Section~\ref{sec:linear-cap}}
\lau{01-09-2017: For now, I will add the linear capabilities such that a linear capability stays linear. It does, however, seem like it would be okay to have linear capabilities become non-linear (I at least don't see anything that would break down completely). One draw back would be that one could ``ruin'' the stack capability by making it non-linear.}
\dominique{7-9-2017: Err... you mean that we could allow non-linear capabilities to be made linear, right?  We definitely do not want to allow the stack or return capability to be made non-linear, as we want to prevent it from being aliased.}
\lau{07-09-2017: I actually did mean linear to non-linear (non-linear to linear does not make sense because the non-linear capability may have aliases). The operation that makes a linear capability non-linear would have to clear said capability. If a linear capability has been made non-linear, then someone who expects a linear capability can observe this difference and behave accordingly, i.e., fail the execution. It will, however, remain an invariant that there are no aliases for linear capabilities.}
\lau{07-09-2017: I had forgotten that return capabilities probably need to be linear. This would call for a ``load from offset'' operation to be able to load an in code stored seal (see the figure I drew about linking/seals).}

@@

\noindent\textbf{gettype}\\
In the definitions of the semantics below, we use a function $\encType{} : \Word \rightarrow \ints$. This is an encoding function for which the specific implementation does not matter. As the words of the two machines differ, we really need two functions which we call $\encType{}_{\var{src}}$ and $\encType{}_{\var{trg}}$. These two functions need to be related in the following way:
% Lau: 19-12-2017: Resolved the belove by adding the conditions.
%\lau{14-09-2017: Insert the relation. My impression is that the the words specific to the source machine should have the same "type" as their implementation at the source level (I am not sure this is the case - maybe it is okay to just relate the values returned by the special words to the ones returned when used with the implementation). }
%\dominique{20-9-2017: I also think gettype for source-specific words should produce the same result than gettype for their implementation.}
\begin{itemize}
\item $\encType{}(((\_,\_),\_,\_,\_))$, $\encType{}(\seal{\_,\_,\_})$, $\encType{}(\sealed{\_,\_})$, and $\encType{}(i)$ where $i\in\ints$ are all destinct.
\item For all $w \in \SealableCaps$ (only the words on the target machine), $\encType{}_{\var{trg}}(w) = \encType{}_{\var{src}}(w)$.
\item Finally, 
\[
\encType{}_{\var{src}}(\src{\stkptr{\_,\_,\_,\_}}) = \encType{}_{\var{trg}}(((\_,\_),\_,\_,\_))
\]
 and 
\[
\encType{}_{\var{src}}(\src{\retptrd(\_,(\_,\_))}) = \encType{}_{\var{src}}(\src{\retptrc(\_,(\_,\_,\_,\_,\_))}) = \encType{}_{\var{trg}}(\sealed{\_,\_})
\]
\end{itemize}
In English this means that each type of word is represented by a distinct value and that the tokens on the source machine has the type of the capability they represent on the target machine.
\begin{align*}
  \sem{\tisptr{r_1}{r_2}}(\Phi) = & \; \updPcAddr{}(\Phi\updReg{r_1}{\encType{\Phi(r_2)}})
\end{align*}

@@

\begin{align*}
  \sem{\tgetlin{r_1}{r_2}}(\Phi) = & 
                                   \begin{cases}
                                     \updPcAddr{\Phi\updReg{r_1}{\linear}} & 
                                     \arraycolsep=0pt
                                     \begin{array}[t]{l}
                                       \isLinear{\Phi(r_2)}
                                     \end{array} \\
                                     \updPcAddr{\Phi\updReg{r_1}{\normal}} & \nonLinear{\Phi(r_2)}
                                   \end{cases}
\end{align*}
\lau{19-12-2017: Resolved the comments below. Seals are all normal and we allow linearity of sealed capabilities to be retrieved as it is observable anyway.}
\lau{15-09-2017: Should getl do something for seals?}
\dominique{4-12-2017: doesn't matter, I think.}
\lau{15-09-2017: Do we want to allow getl and getp to work on sealed capabilities?}
\dominique{Linearity of sealed caps is already observable anyway, so no point in hiding it.  Permissions of the sealed capability should be hidden, I think, mostly out of conservativity (hide as much as possible!).}\\
\dominique{4-12-2017: Couldn't the definitions above use $\pi_b$, $\pi_e$, $\pi_l$ and $\pi_a$ and $\pi_p$ (the latter two to be added) ?}
\lau{12-12-2017: $\pi_\_$ also projects from sealed capabilities, so the case of sealed capabilities would have to be explicitely excluded (with getl as the exception).}

@@

\noindent\textbf{xjmp}\\
\lau{22-09-2017: Can we allow the callee to return any stack pointer? At the moment, it is not possible to create an empty stack pointer which may be a saving grace. If we at some point change this, so you can split capabilities into empty ones, then we need to consider whether it is safe for callees to return empty stacks (I think it is okay as long as we always have some stack usage).}
\lau{28-09-2017: For now, the jump to a return pointer does the stack end point check. We could possibly move this to the proposed semantic condition as well.}
\begin{align*}
  \sem{\txjmp{r_1}{r_2}}(\Phi) = &
                                   \begin{cases}
                                     \arraycolsep=0pt
                                     \begin{array}[t]{rl}
                                       \Phi & \updReg{r_1}{w_1}\\
                                            & \updReg{r_2}{w_2}\\
                                            & \updReg{\pcreg}{c_\code}\\
                                            & \updReg{\rdata}{c_\data}
                                     \end{array} &
                                     \begin{array}[t]{l}
                                       \Phi(r_1) = \sealed{\sigma_1,c_\code} \tand \\
                                       \Phi(r_2) = \sealed{\sigma_2,c_\data} \tand \\
                                       \sigma_1 = \sigma_2 \tand \\
                                       \nonExec{c_\data} \tand \\
                                       w_1 = \linCons{\Phi(r_1)} \tand \\
                                       w_2 = \linCons{\Phi(r_2)}
                                     \end{array}
                                     \\&\\
                                     \sourcecolor\left.
                                       \arraycolsep=0pt
                                       \begin{array}[t]{rl}
                                         \Phi' & \updReg{r_2}{0} \\
                                               & \updReg{\pcreg}{\opc}\\
                                               & \updReg{\rdata}{0}\\
                                               & \updReg{\rstk}{c_\stk}
                                       \end{array}\right.&
                                     \sourcecolor\left.
                                       \begin{array}[t]{l}
                                         \Phi(r_1) = \retptrc(\sigma_1,\_) \tand \\
                                         \Phi(r_2) = \retptrd(\sigma_2,(\aaddr_\stk,\_)) \tand \\
                                         \sigma_1 = \sigma_2 \tand\\
                                         \Phi(\rstk) = \stkptr{\rw,\stkb,\eaddr_\stk,\_} \tand \\
                                         \Phi = (\mem,\reg,\stkf::\stk,\ms_\stk) \tand \\
                                         \stkf = (\opc,\ms_{\stk,\priv}) \tand \\
                                         \dom(\ms_{\stk,\priv}) = [\eaddr_\stk+1,\eaddr_{\stk,\priv}] \tand\\
                                         c_\stk = \stkptr{\rw,\stkb,\eaddr_{\stk,\priv},\aaddr_\stk} \tand\\
                                         \Phi' = (\mem,\reg,\stk,\ms_{\stk,\priv} \uplus \ms_\stk) 
                                       \end{array}\right.
                                     \\
                                     \failed & \totherwise
                                   \end{cases}
\end{align*}

\noindent\textbf{cseal}\\
\begin{align*}
  \sem{\tcseal{r_1}{r_2}{r_3}}(\Phi) = &
                                  \begin{cases}
                                    \updPcAddr{}\left(
                                    \arraycolsep=0pt
                                    \begin{array}{rl}
                                      \Phi&\updReg{r_1}{\vsc}\\
                                          &\updReg{r_2}{w_2}
                                    \end{array}\right)
&
                                    \arraycolsep=0pt
                                    \begin{array}[t]{l}
                                      \Phi(r_2) \in \SealableCaps \tand \\
                                      \Phi(r_3) = \seal{\sigma_\baddr, \sigma_\eaddr,\sigma} \tand \\
                                      \withinBounds{\Phi(r_3)} \tand \\
                                      \vsc = \sealed{\sigma,\Phi(r_2)} \tand\\
                                      w_2 = \linCons{\Phi(r_2)}                                     
                                    \end{array}
                                    \\
                                    \failed & \totherwise
                                  \end{cases}
\end{align*}

@@

@@

@@

                                   c_2 = \seal{n+1,\sigma_\eaddr,\sigma} \tand \\

@@

@@

                                  c = \seal{\sigma_{\baddr,2},\sigma_{\eaddr,3},\sigma} \tand \\

@@

                                                   \stk' = (\opc,\ms_{\stk,\priv},\aaddr_\stk) :: \stk \tand\\

@@

Eventhough we call them linear, they are more like \emph{affine} capabilities because our current operational semantics allows for them to be overwritten. We choose to stick with the name as the term linear is also being misused by others for instance in Rust.

>>>>

  \caption{Permission hierarchy}
  \label{fig:perm-hier}
\end{figure}
\lau{13-09-2017: Note that most of the discussion below is now captured in Section~\ref{sec:linear-cap}}
\lau{01-09-2017: For now, I will add the linear capabilities such that a linear capability stays linear. It does, however, seem like it would be okay to have linear capabilities become non-linear (I at least don't see anything that would break down completely). One draw back would be that one could ``ruin'' the stack capability by making it non-linear.}
\dominique{7-9-2017: Err... you mean that we could allow non-linear capabilities to be made linear, right?  We definitely do not want to allow the stack or return capability to be made non-linear, as we want to prevent it from being aliased.}
\lau{07-09-2017: I actually did mean linear to non-linear (non-linear to linear does not make sense because the non-linear capability may have aliases). The operation that makes a linear capability non-linear would have to clear said capability. If a linear capability has been made non-linear, then someone who expects a linear capability can observe this difference and behave accordingly, i.e., fail the execution. It will, however, remain an invariant that there are no aliases for linear capabilities.}
\lau{07-09-2017: I had forgotten that return capabilities probably need to be linear. This would call for a ``load from offset'' operation to be able to load an in code stored seal (see the figure I drew about linking/seals).}

@@

\noindent\textbf{gettype}\\
In the definitions of the semantics below, we use a function $\encType{} : \Word \rightarrow \ints$. This is an encoding function for which the specific implementation does not matter. As the words of the two machines differ, we really need two functions which we call $\encType{}_{\var{src}}$ and $\encType{}_{\var{trg}}$. These two functions need to be related in the following way:
% Lau: 19-12-2017: Resolved the belove by adding the conditions.
%\lau{14-09-2017: Insert the relation. My impression is that the the words specific to the source machine should have the same "type" as their implementation at the source level (I am not sure this is the case - maybe it is okay to just relate the values returned by the special words to the ones returned when used with the implementation). }
%\dominique{20-9-2017: I also think gettype for source-specific words should produce the same result than gettype for their implementation.}
\begin{itemize}
\item $\encType{}(((\_,\_),\_,\_,\_))$, $\encType{}(\seal{\_,\_,\_})$, $\encType{}(\sealed{\_,\_})$, and $\encType{}(i)$ where $i\in\ints$ are all distinct.
\item For all $w \in \SealableCaps$ (only the words on the target machine), $\encType{}_{\var{trg}}(w) = \encType{}_{\var{src}}(w)$.
\item Finally, 
\[
\encType{}_{\var{src}}(\src{\stkptr{\_,\_,\_,\_}}) = \encType{}_{\var{trg}}(((\_,\_),\_,\_,\_))
\]
 and 
\[
\encType{}_{\var{src}}(\src{\retptrd(\_,(\_,\_))}) = \encType{}_{\var{src}}(\src{\retptrc(\_,(\_,\_,\_,\_,\_))}) = \encType{}_{\var{trg}}(\sealed{\_,\_})
\]
\end{itemize}
In English this means that each type of word is represented by a distinct value and that the tokens on the source machine has the type of the capability they represent on the target machine.
\begin{align*}
  \sem{\tisptr{r_1}{r_2}}(\Phi) = & \; \updPcAddr{}(\Phi\updReg{r_1}{\encType{\Phi(r_2)}})
\end{align*}

@@

\begin{align*}
  \sem{\tgetlin{r_1}{r_2}}(\Phi) = & 
                                   \begin{cases}
                                     \updPcAddr{\Phi\updReg{r_1}{\linear}} & 
                                     \arraycolsep=0pt
                                     \begin{array}[t]{l}
                                       \isLinear{\Phi(r_2)}
                                     \end{array} \\
                                     \updPcAddr{\Phi\updReg{r_1}{\normal}} & \nonLinear{\Phi(r_2)}
                                   \end{cases}
\end{align*}
\lau{19-12-2017: Resolved the comments below. Seals are all normal and we allow linearity of sealed capabilities to be retrieved as it is observable anyway.}
\lau{15-09-2017: Should getl do something for seals?}
\dominique{4-12-2017: doesn't matter, I think.}
\lau{15-09-2017: Do we want to allow getl and getp to work on sealed capabilities?}
\dominique{Linearity of sealed caps is already observable anyway, so no point in hiding it.  Permissions of the sealed capability should be hidden, I think, mostly out of conservativity (hide as much as possible!).}\\
\dominique{4-12-2017: Couldn't the definitions above use $\pi_b$, $\pi_e$, $\pi_l$ and $\pi_a$ and $\pi_p$ (the latter two to be added) ?}
\lau{12-12-2017: $\pi_\_$ also projects from sealed capabilities, so the case of sealed capabilities would have to be explicitly excluded (with getl as the exception).}

@@

\noindent\textbf{xjmp}\\
\lau{22-09-2017: Can we allow the callee to return any stack pointer? At the moment, it is not possible to create an empty stack pointer which may be a saving grace. If we at some point change this, so you can split capabilities into empty ones, then we need to consider whether it is safe for callees to return empty stacks (I think it is okay as long as we always have some stack usage).}
\lau{28-09-2017: For now, the jump to a return pointer does the stack end point check. We could possibly move this to the proposed semantic condition as well.}
\begin{align*}
  \sem{\txjmp{r_1}{r_2}}(\Phi) = &
                                   \begin{cases}
                                     \arraycolsep=0pt
                                     \begin{array}[t]{rl}
                                       \Phi & \updReg{r_1}{w_1}\\
                                            & \updReg{r_2}{w_2}\\
                                            & \updReg{\pcreg}{c_\code}\\
                                            & \updReg{\rdata}{c_\data}
                                     \end{array} &
                                     \begin{array}[t]{l}
                                       \Phi(r_1) = \sealed{\sigma_1,c_\code} \tand \\
                                       \Phi(r_2) = \sealed{\sigma_2,c_\data} \tand \\
                                       \sigma_1 = \sigma_2 \tand \\
                                       \nonExec{c_\data} \tand \\
                                       w_1 = \linCons{\Phi(r_1)} \tand \\
                                       w_2 = \linCons{\Phi(r_2)}
                                     \end{array}
                                     \\&\\
                                     \sourcecolor\left.
                                       \arraycolsep=0pt
                                       \begin{array}[t]{rl}
                                         \Phi' & \updReg{r_2}{0} \\
                                               & \updReg{\pcreg}{\opc}\\
                                               & \updReg{\rdata}{0}\\
                                               & \updReg{\rstk}{c_\stk}\\
                                               & \updReg{\rtmp{1}}{0}\\
                                               & \updReg{\rtmp{2}}{0}
                                       \end{array}\right.&
                                     \sourcecolor\left.
                                       \begin{array}[t]{l}
                                         \Phi(r_1) = \retptrc(\sigma_1,\_) \tand \\
                                         \Phi(r_2) = \retptrd(\sigma_2,(\aaddr_\stk,\_)) \tand \\
                                         \sigma_1 = \sigma_2 \tand\\
                                         \Phi(\rstk) = \stkptr{\rw,\stkb,\eaddr_\stk,\_} \tand \\
                                         \Phi = (\mem,\reg,\stkf::\stk,\ms_\stk) \tand \\
                                         \stkf = (\opc,\ms_{\stk,\priv}) \tand \\
                                         \dom(\ms_{\stk,\priv}) = [\eaddr_\stk+1,\eaddr_{\stk,\priv}] \tand\\
                                         c_\stk = \stkptr{\rw,\stkb,\eaddr_{\stk,\priv},\aaddr_\stk} \tand\\
                                         \Phi' = (\mem,\reg,\stk,\ms_{\stk,\priv} \uplus \ms_\stk) 
                                       \end{array}\right.
                                     \\
                                     \failed & \totherwise
                                   \end{cases}
\end{align*}

\noindent\textbf{cseal}\\
\begin{align*}
  \sem{\tcseal{r_1}{r_2}{r_3}}(\Phi) = &
                                  \begin{cases}
                                    \updPcAddr{}\left(
                                    \arraycolsep=0pt
                                    \begin{array}{rl}
                                      \Phi&\updReg{r_2}{w_2}\\
                                          &\updReg{r_1}{\vsc}
                                    \end{array}\right)
&
                                    \arraycolsep=0pt
                                    \begin{array}[t]{l}
                                      \Phi(r_2) \in \SealableCaps \tand \\
                                      \Phi(r_3) = \seal{\sigma_\baddr, \sigma_\eaddr,\sigma} \tand \\
                                      \withinBounds{\Phi(r_3)} \tand \\
                                      \vsc = \sealed{\sigma,\Phi(r_2)} \tand\\
                                      w_2 = \linCons{\Phi(r_2)}                                     
                                    \end{array}
                                    \\
                                    \failed & \totherwise
                                  \end{cases}
\end{align*}

@@

%Alternative way to phrase the same thing: take the set of addresses governed by a linear capability. After a split, the union of the two sets that consist of the addresses the two capabilities govern should be equal to the set of addresses the original capability governed. Further, the two sets of addresses two capabilities uptained from a split govern should be disjoint.

@@

                                   \Phi&\updReg{r_1}{c_1}\\

@@

                                   c_2 = \seal{n+1,\sigma_\eaddr,\sigma}         

@@

                                  \Phi&\updReg{r_1}{c}

@@

                                  c = \seal{\sigma_{\baddr,2},\sigma_{\eaddr,3},\sigma}

@@

                                                   \stk' = (\opc,\ms_{\stk,\priv}) :: \stk \tand\\

@@

Even though we call them linear, they are more like \emph{affine} capabilities because our current operational semantics allows for them to be overwritten. We choose to stick with the name as the term linear is also being misused by others for instance in Rust.
